The locally Gaussian partial correlation

### 2020-09-16 lgpc_JBES.Rmd
A few minor details in the main paper by Håkon

Equations \@ref(eq:matrixlocalpartial), \@ref(eq:definition) and their empirical counterparts \@ref(eq:localpartialcov) and \@ref(eq:estimate) demonstrate clearly that the LGPC is nothing more than a deterministic function of the local correlation matrix, in the same way as the ordinary partial correlation is a function of the ordinary correlation matrix. The asymptotic behavior of $\halpha(\z)$ can thus be derived directly from the asymptotic behavior of $\hR(\z)$, which, for the most part, has been established in earlier works, see e.g. @tjostheim2013local and Otneim and Tjøstheim [-@otneim2017locally; -@otneim2017conditional]. Some key results follow below, and we refer to the supplementary material, Sections B--D for proofs of the asymptotic distributional results in this subsection, and to Section A.3 for a discussion of more general situations and for a result on consistency.

@@

Then, by letting $b_n^2$ (assuming identical convergence rates) mean the product $b_{nj}b_{nk}$ corresponding to the pairs of variables defining the components in $\frho$, -->
\begin{equation}
\sqrt{nb_n^2}\left(\hfrho(\z) - \frho_0(\z)\right) \stackrel{d}{\rightarrow} \mathcal{N}(\bm{0}, \fLambda),
\label{eq:ot2018}
\end{equation}
<!--
where $\fOmega$ is a diagonal matrix with components
\begin{equation}
\fOmega^{(\ell,\ell)} = 
\frac{f_{\ell}(\z_{\ell})\int K^2(\y_{\ell})\,\textrm{d}\y_{\ell}}
{u^2(\z_{\ell},\rho_{0,\ell}(\z_{\ell}))}
\psi^2(\z_{\ell}, \rho_{0,\ell}(\z_{\ell})),
(\#eq:asymcov)
\end{equation}
where $\ell = 1,\ldots,p(p-1/2)$ runs over all pairs of variables, and $f_{\ell}(\cdot)$ is the bivariate marginal density function of the pair $\Z_{\ell}$.
```

@@

\begin{equation}
\sqrt{nb_n^m}\left(\halpha(\z) - \alpha(\z)\right) \stackrel{d}{\rightarrow} \mathcal{N}\left(0, \nabla g(\frho)^T\fLambda\nabla g(\frho)\right),
(\#eq:limitlgpc)
\end{equation}
where $m$ is equal to 2 or 5, and $\fLambda$ is either equal to the leading term of $\Jb^{-1}\Mb(\Jb^{-1})^T$, or $\fLambda$ in \@ref(eq:ot2018), depending on whether we use the full trivariate locally Gaussian fit described in Section \@ref(chap:trivariate-full) or the sequentially pairwise simplification described in Section \@ref(chap:bivariate). In any case, it is only a matter of basic differentiation to work out an expression for $\nabla g(\cdot)$, and this task has been deferred to Section C of the online supplement in the case where $d_1 = d_2 = 1$.

>>>>

Equations \@ref(eq:matrixlocalpartial), \@ref(eq:definition) and their empirical counterparts \@ref(eq:localpartialcov) and \@ref(eq:estimate) demonstrate clearly that the LGPC is nothing more than a deterministic function of the local correlation matrix, in the same way as the ordinary partial correlation is a function of the ordinary correlation matrix. The asymptotic behavior of $\halpha(\z)$ can thus be derived directly from the asymptotic behavior of $\hR(\z)$, which, for the most part, has been established in earlier works, see e.g. @tjostheim2013local and Otneim and Tjøstheim [-@otneim2017locally; -@otneim2017conditional]. Some key results follow below, and we refer to the supplementary material, Sections B--D, for proofs of the asymptotic distributional results in this subsection, and to Section A.3 for a discussion of more general situations and for a result on consistency.

@@

Then, by letting $b_n^2$ (assuming identical convergence rates) mean the product $b_{nj}b_{nk}$ corresponding to the pairs of variables defining the components in $\frho$, -->
\begin{equation}
\sqrt{nb^2}\left(\hfrho(\z) - \frho_0(\z)\right) \stackrel{d}{\rightarrow} \mathcal{N}(\bm{0}, \fLambda),
\label{eq:ot2018}
\end{equation}
<!--
where $\fOmega$ is a diagonal matrix with components
\begin{equation}
\fOmega^{(\ell,\ell)} = 
\frac{f_{\ell}(\z_{\ell})\int K^2(\y_{\ell})\,\textrm{d}\y_{\ell}}
{u^2(\z_{\ell},\rho_{0,\ell}(\z_{\ell}))}
\psi^2(\z_{\ell}, \rho_{0,\ell}(\z_{\ell})),
(\#eq:asymcov)
\end{equation}
where $\ell = 1,\ldots,p(p-1/2)$ runs over all pairs of variables, and $f_{\ell}(\cdot)$ is the bivariate marginal density function of the pair $\Z_{\ell}$.
```

@@

\begin{equation}
\sqrt{nb^m}\left(\halpha(\z) - \alpha(\z)\right) \stackrel{d}{\rightarrow} \mathcal{N}\left(0, \nabla g(\frho)^T\fLambda\nabla g(\frho)\right),
(\#eq:limitlgpc)
\end{equation}
where $m$ is equal to 2 or 5, and $\fLambda$ is either equal to the leading term of $\Jb^{-1}\Mb(\Jb^{-1})^T$, or $\fLambda$ in \@ref(eq:ot2018), depending on whether we use the full trivariate locally Gaussian fit described in Section \@ref(chap:trivariate-full) or the sequentially pairwise simplification described in Section \@ref(chap:bivariate). In any case, it is only a matter of basic differentiation to work out an expression for $\nabla g(\cdot)$, and this task has been deferred to Section C of the online supplement in the case where $d_1 = d_2 = 1$.

### 2020-09-16 lgpc_JBES.tex
A few minor details in the main paper by Håkon

Equations \eqref{eq:matrixlocalpartial}, \eqref{eq:definition} and their empirical counterparts \eqref{eq:localpartialcov} and \eqref{eq:estimate} demonstrate clearly that the LGPC is nothing more than a deterministic function of the local correlation matrix, in the same way as the ordinary partial correlation is a function of the ordinary correlation matrix. The asymptotic behavior of \(\halpha(\z)\) can thus be derived directly from the asymptotic behavior of \(\hR(\z)\), which, for the most part, has been established in earlier works, see e.g.~Tjøstheim and Hufthammer (\protect\hyperlink{ref-tjostheim2013local}{2013}) and Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017locally}{2017}, \protect\hyperlink{ref-otneim2017conditional}{2018}). Some key results follow below, and we refer to the supplementary material, Sections B--D for proofs of the asymptotic distributional results in this subsection, and to Section A.3 for a discussion of more general situations and for a result on consistency.

@@

The pairwise estimation structure that we describe in Section \ref{chap:bivariate} simplifies the asymptotic analysis of \(\hR(\z)\) considerably. The result is essentially available in the literature as Theorem 2 in Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017conditional}{2018}), who state under similar assumptions as in Theorem \ref{thm:loccor3} that
\begin{equation}
\sqrt{nb_n^2}\left(\hfrho(\z) - \frho_0(\z)\right) \stackrel{d}{\rightarrow} \mathcal{N}(\bm{0}, \fLambda),
\label{eq:ot2018}
\end{equation}
where the asymptotic covariance matrix \(\fLambda\) is defined as \(\fSigma\) in Theorem 2 of Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017conditional}{2018}). There are two important differences compared to Theorem \ref{thm:loccor3}, however. First of all, we have that the convergence rate in the simplified pairwise case is equal to the usual nonparametric rate of \(1/(nb^2)^{1/2}\). This is because we estimate the local correlations one by one based on pairs of variables. The second difference is that the leading term of the asymptotic covariance matrix in Theorem 2 by Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017conditional}{2018}) has a simple analytic expression that can be used to construct, for instance, confidence intervals. We refer to Section A.3 in the online supplement for further details regarding the limiting distribution of the LGPC as well conditions for consistency of the estimates.

@@

\begin{equation}
\sqrt{nb_n^m}\left(\halpha(\z) - \alpha(\z)\right) \stackrel{d}{\rightarrow} \mathcal{N}\left(0, \nabla g(\frho)^T\fLambda\nabla g(\frho)\right),
\label{eq:limitlgpc}
\end{equation}
where \(m\) is equal to 2 or 5, and \(\fLambda\) is either equal to the leading term of \(\Jb^{-1}\Mb(\Jb^{-1})^T\), or \(\fLambda\) in \eqref{eq:ot2018}, depending on whether we use the full trivariate locally Gaussian fit described in Section \ref{chap:trivariate-full} or the sequentially pairwise simplification described in Section \ref{chap:bivariate}. In any case, it is only a matter of basic differentiation to work out an expression for \(\nabla g(\cdot)\), and this task has been deferred to Section C of the online supplement in the case where \(d_1 = d_2 = 1\).

>>>>

Equations \eqref{eq:matrixlocalpartial}, \eqref{eq:definition} and their empirical counterparts \eqref{eq:localpartialcov} and \eqref{eq:estimate} demonstrate clearly that the LGPC is nothing more than a deterministic function of the local correlation matrix, in the same way as the ordinary partial correlation is a function of the ordinary correlation matrix. The asymptotic behavior of \(\halpha(\z)\) can thus be derived directly from the asymptotic behavior of \(\hR(\z)\), which, for the most part, has been established in earlier works, see e.g.~Tjøstheim and Hufthammer (\protect\hyperlink{ref-tjostheim2013local}{2013}) and Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017locally}{2017}, \protect\hyperlink{ref-otneim2017conditional}{2018}). Some key results follow below, and we refer to the supplementary material, Sections B--D, for proofs of the asymptotic distributional results in this subsection, and to Section A.3 for a discussion of more general situations and for a result on consistency.

@@

The pairwise estimation structure that we describe in Section \ref{chap:bivariate} simplifies the asymptotic analysis of \(\hR(\z)\) considerably. The result is essentially available in the literature as Theorem 2 in Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017conditional}{2018}), who state under similar assumptions as in Theorem \ref{thm:loccor3} that
\begin{equation}
\sqrt{nb^2}\left(\hfrho(\z) - \frho_0(\z)\right) \stackrel{d}{\rightarrow} \mathcal{N}(\bm{0}, \fLambda),
\label{eq:ot2018}
\end{equation}
where the asymptotic covariance matrix \(\fLambda\) is defined as \(\fSigma\) in Theorem 2 of Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017conditional}{2018}). There are two important differences compared to Theorem \ref{thm:loccor3}, however. First of all, we have that the convergence rate in the simplified pairwise case is equal to the usual nonparametric rate of \(1/(nb^2)^{1/2}\). This is because we estimate the local correlations one by one based on pairs of variables. The second difference is that the leading term of the asymptotic covariance matrix in Theorem 2 by Otneim and Tjøstheim (\protect\hyperlink{ref-otneim2017conditional}{2018}) has a simple analytic expression that can be used to construct, for instance, confidence intervals. We refer to Section A.3 in the online supplement for further details regarding the limiting distribution of the LGPC as well conditions for consistency of the estimates.

@@

\begin{equation}
\sqrt{nb^m}\left(\halpha(\z) - \alpha(\z)\right) \stackrel{d}{\rightarrow} \mathcal{N}\left(0, \nabla g(\frho)^T\fLambda\nabla g(\frho)\right),
\label{eq:limitlgpc}
\end{equation}
where \(m\) is equal to 2 or 5, and \(\fLambda\) is either equal to the leading term of \(\Jb^{-1}\Mb(\Jb^{-1})^T\), or \(\fLambda\) in \eqref{eq:ot2018}, depending on whether we use the full trivariate locally Gaussian fit described in Section \ref{chap:trivariate-full} or the sequentially pairwise simplification described in Section \ref{chap:bivariate}. In any case, it is only a matter of basic differentiation to work out an expression for \(\nabla g(\cdot)\), and this task has been deferred to Section C of the online supplement in the case where \(d_1 = d_2 = 1\).
